---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
#  ,out.width = "60%"
```
<img src = "man/figures/spca_Logo_bordered.png" align = "left" height = "110" alt = "spca logo" />

# Package spca

<!-- badges: start -->
[![R-CMD-check](https://github.com/merolagio/spca/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/merolagio/spca/actions/workflows/R-CMD-check.yaml)
[![License](https://img.shields.io/badge/license-MIT-green.svg)](LICENSE)

[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html)
[![Project Status: Active](https://img.shields.io/badge/status-active-success.svg)]()
<!-- badges: end -->

This package contains functions to compute, print and plot Least Squares Sparse Principal Components Analysis (LS-SPCA). Methodological details and full presentation can be found in the extended_vignette document.

The  main function *spca()* computes the sparse loadings and various statistics, such as the variance explained by each sparse component (sPC). print, summery and plot methods are available. 

Utilities available are `compare_spca` (to compare two or more spca solutions), `aggregate_by_scale` (to visualize the contribution by scale) and `new.spca` (to create an `spca' object from a set of loadings).

## Installation

You can install the development version of spca from [GitHub](https://github.com/) with:

``` r
# `remotes' is the lightest alternative
install.packages("remotes")
remotes::install_github("merolagio/spca")
#or
#install.packages("devtools")
devtools::install_github("merolagio/spca")
# or
# install.packages("pak")
pak::pak("merolagio/spca")
```

## Example

### Load data
The `holzinger` dataset is the small classic Holzinger-Swineford dataset with 145 cases on 12 variables grouped in 4 scales.
```{r load_data, results = 'hide', message = FALSE, warning = FALSE}
library(spca)
data(holzinger)
```

### Preliminary PCA
```{r pca_checks, message = FALSE, warning = FALSE, fig.show="hold", out.width="47%", fig.width=4, fig.height=4}
ho_pca = pca(holzinger, screeplot =  TRUE, qq_plot = TRUE)
summary(ho_pca, cols = 10)
```
We settle for 4 components

### Compute the sparse loadings
Important parameters in the `spca` function are: $\alpha$ the minimum $R^2$ [default]) or the minimum proportion of cumulative vexp of the PCs reproduced by the sPCs; *n_comps* the number of components to compute; *method* the LS-SPCA method to use ("u" for uncorrelated, "c" for correlated [default]) and "p" for projection; and *var_selection* ("forward" [default], "stepwise", or "backward"). See the `spca` help for details on these parameters and more.

The following command computes four sPCs with default seetings: *alpha = 0.95*, *forward* variable selection with the *cSPCA* method. We expect each sPC yield at least 0.95% cumulative VEXP, allowing some very mild correlation between sPCs.
```{r run_spca, message = FALSE, warning = FALSE}
myspca = spca(holzinger, n_comps = 4)
```

### Inspect spca results
Methods are *print*, *plot* (several options available) and *summary*.
By defaut, plot and print show the percentage *contributions*, that is the loadings scaled to have sum of their absolute values equal to 1.
```{r methods, message = TRUE, warning = FALSE, out.width="60%"}
myspca # print

summary(myspca, cor_with_pc = TRUE)

plot(myspca, plot_type = "bar")

#sPCs correlation
round(myspca$spc_cor, 2)

```

Other types of plots are available.

Circular, 
```{r circular, message = FALSE, warning = FALSE, fig.width = 5, fig.height = 3}
plot(myspca, plot_type = "c") # "c" for "circular"
```

Heatmap
```{r heatmap, message = FALSE, warning = FALSE, fig.width = 5, fig.height = 3}
plot(myspca, plot_type = "h", controls = list(legend_position = "b")) # "h" is enough to call "heatmap" type and "b" to indicate "bottom".
```



## variable groups
The variables in the `holzinger` dataset belong to four different scales, recorded in the factor `holzinger_scales`. These can be differenciated in the barplot
```{r groups, message = FALSE, warning = FALSE, fig.width = 5, fig.height = 3}
plot(myspca, plot_type = "bars", variable_groups = holzinger_scales, controls = list(legend_position = "right")) 

aggregate_by_group(myspca,groups = holzinger_scales)
```

## comparison of two or more spca solutions
Compare the *CSPCA* solutions with *alpha = 0.95* those with  *alpha = 0.90*.
```{r spca90, message = FALSE, warning = FALSE}
myspca90 = spca(holzinger, n_comps = 4, alpha = 0.9)

compare_spca(obj_list = list(myspca, myspca90), methods_names = c("alpha = 95", "alpha = 90"))
```
