---
title: "Introduction to the spca package"
author: "Giovanni Merola<br>
RMIT International University Vietnam<br>
email: lsspca@gmail.com<br>
repository: https://github.com/merolagio/spca"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  rmarkdown::html_document:
    toc: true
    theme: united
    highlight: haddock
    fig_caption: true
bibliography: spca.bibtex    
vignette: >
  %\VignetteIndexEntry{Introduction to spca}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}


---

```{r, echo = FALSE, message = FALSE}
library(spca)
library(formatR)
library(knitr)
knitr::opts_chunk$set(
  comment = "#>",
  error = FALSE,
  tidy = FALSE)
options(width=120)
```

### Foreword 
Unfortunately, creating package vignettes in LaTex using Rmarkdown is difficult because of the fragilty of the interface. After spending several days on this I finally gave in and rewrote them in markdown. This was painful, as I had to spend a lot of time on rewriting the documentation instead of debugging the package.  This means that 
the quality of these vignettes is not what I would have liked. On the positive side, I can now send cool emails with markdown :-)

### Intro
`spca` is an R package for running Sparse Principal Component Analysis. It implements the **LS SPCA** approach that computes the Least Squares estimates of sparse PCs (@mer). Unlike other SPCA methods, these solutions maximise the variance of the data explained. 

The implementation is completely in `R`, so it can only be run on *small* datasets (the limit depends on the hardware used, we were able to solve problems with about 1000 variables in minutes). The package is self contained as it only depends on the library `MASS` which is part of the basic distribution of R. 

Details about LS SPCA and the methodology impmented in the package can be found in ([Merola, 2014. arXiv](http://arxiv.org/abs/1406.1381 "Pre-print")) and in the forthcoming peer reviewed paper. 

I had difficulties publishing the LS SPCA paper, possibly because LS SPCA improves on existing methods. This is confirmed by the fact that Technometrics' chief editor, Dr Qiu, rejected the paper endorsing a report stating that: **the LS criterion is a new measure used ad-hoc  :-D** This on top of a number of blatantly wrong arguments. Dr Qiu added in his rejection letter that the algorithm wasn't scalable. Now, this is arbitrary because computational efficiency is not in the scope and aims of Technometrics, A reviewer from *ANZJS* asked me to **compare the about 20 existing SPCA methods with mine on more datasets** (only because I show that my solutions maximise the variance explained and theirs don't)! However, the editors of this journal accepted my refusal to do so.

### A little math
Principal Component Analysis was developed by @pea to attain the components that minimised the LS criterion when approximating the data matrix. If $X$ is a matrix with *n* rows of observations on *p* variables, the PCs are defined by the loadings $a_j$ as $t_j = X a_j$. The matrix of $d<p$ PCs, $T = XA$, is derived as the regressors that minimise the Residual Sum of Squares. By the principle of the Extra Sum of Squares the components can be constrained to be uncorrelated without loss of optimality. Therefore the PCA problem is obtained by solving: 
$$
A = \text{arg}\min ||X - TP'||^2 = \text{arg}\max\frac{A'SSA}{A'SA} = \text{arg}\max \sum_{1}^d \frac{a_j'SSa_j}{a_j'Sa_j}\\
\text{subject to}\ a_j'Sa_k = 0,\ j\neq k,
$$
where $S$ is the covariance matrix of the $x$ variables. The terms in the last summation are the *variance explained* by each component. The solutions are proportional to the eigenvectors of $S$ corresponding to the eigenvalues taken in nonicreasing order. It is well known that the eigenvectors are mutually uncorrelated.

@hot gives the PCs' loadings as the eigenvectors of $S$ with unit Euclidean norm. Using this normalisation the maximisation of the variance explained by each component simplifies to
$$
A =  \text{arg}\max \sum_{1}^d {a_j'Sa_j}\\
\text{subject to}\ a_j'Sa_k = \delta_{jk},
$$
where $\delta_{jk} = 1$ if $j=k$ and  $\delta_{jk} = 0$ otherwise.

Because of its simplicity, Hotelling's derivation has been adopted for popularizing PCA among pratictioners. This choice was unfortunate because the original objective of minimising the LS criterion has been somewhat forgotten. However, other than in Person's original paper, the LS derivation is given in several books and papers cited in my paper ( *e.g.* @ten and @ize08).  

When cardinality constraints (also called $L_0$ *norm* constraints) are imposed on the original PCA problem, the loadings are no longer eigenvectors of $S$. Therefore, Hotelling's simplification is no longer equivalent to the variance explained. Furthermore, by the Cauchy-Schwartz inequality:
$$
\frac{a_j'SSa_j}{a_j'Sa_j} \geq \frac{a_j'Sa_j}{a_j'a_j}
$$
for any square matrix *S*, with equality if and only if the vectors $a_j$ are eigenvectors of $S$. Therefore, the components with maximal variance are suboptimal for explaining the variance. 

Other SPCA methods apply cardinality constraints to Hotelling's definition, hence not optimising the variance explained. Instead, in LS SPCA we derive the loadings from to Pearson's LS optimisation adding cardinality constraints.

The uncorrelated LS SPCA solutions are constrained Reduced Rank Regression solutions (see @ize, for the unconstrained solutions). The uncorrelatedness constraints limit the amount of variance explained by the solutions and require that the loadings have cardinality not smaller than their rank. Even though uncorrelated components are easier to interpret, in some cases uncorrelated ones can be useful. Therefore, we also provide correlated sparse loadings that approximately minimise the LS criterion.

### Optimisation Models
Finding the optimal indices for an *spca* solution is an intractable NP-hard problem.  

Therefore, we find the solutions through two greedy algorithms: Branch-and-Bound (**BB**) and Backward Elimination (**BE**).

* **BB** searches for the solutions that sequentially maximise the variance explained under the constraints. The solutions may not be a global maximum when more than one component is computed. The BB algorithm is a modification of @fra's (thanks!).

* **BE** has the goal of attaining larger contributions while minimising the LS criterium. It sequentially eliminates the smallest contributions (in absolute value) from a non-sparse solution. 

* The  **BE** solutions will generally explain less variance than the **BB** ones. However, the **BE** algorithm is much faster and the solutions, usually, have larger loadings. The algorithm is illustrated in more details in the *BE Algorithm* vignette `vignettes("BE algorithm", package = "spca").


### Use of the package

**SPCA aims to obtain interpretable solutions**

Interpretability is not univocally defined. Hence, for a problem there exist a number of competing solutions. In Factor Analysis literature there is plenty of discussion about the  definition of *interpretable* and *simple* solution (as qualities and mathematical functions). 

* *Simplicity* can be defined by different measures, being linked to sparseness, parsimony, variance explained and size of the loadings. 

* *interpretability* is also linked to which of the variables are included in the solution  and is not measurable.
    * it usually requires expert knowledge.
    
Therefore, for a given problem there usually exist several competing *simple* and *interpretable* solutions. 

`spca` **is implemented as an exploratory data analysis tool** 

The cardinality of the components can be chosen interactively after inspecting trace and plots of solutions of different cardinality.

Solutions can be also computed non-interactively so as to:

* be uncorrelated with the others or not.
* have a minimal cardinality. 
* reproduce a given proportion of the variance explained by the full PCs. 
* have only contributions larger than a given threshold.

**spca** contains methods for plotting and printing the solutions and for comparing different ones. In this way the solution can be chosen with respect to several different characteristicsm which cannot be all included in a function at the same time.  

`spca` can be helpful also in a confirmatory stage of the analysis, in fact
* the components can be constrained to be made up of only a subset of the variables.

### Functions
The workhorse of the package is the function `spca`, which computes the optimal solutions for a given set of indices.

It is called simply with a list of indices and the the flags for correlated or uncorrelated components (one for each component, if necessary)
```{r spca,   comment = "", echo = FALSE}
usage(spca)
```

The functions `spcabb` and `spcabe` implement the **BB** and **BE** searches, respectively.

```{r spcabb, comment = "", echo = FALSE}
usage(spcabb)
```

```{r spcabe, comment = "", echo = FALSE}
usage(spcabe)
```

With`help(spcabb)` and `help(spcabe)` you will find examples of using spca and the utilities. In the `vignettes(spca)` you will find a more complete example and details on the methods. These are available also in the *Manual* and a more complete example is given in the *Advanced Example* vignette.

There is also the function 'pca' which computes the PCA solutions and returns an *spca* object. The function is called as:


```{r pca, comment = "", echo = FALSE}
usage(pca)
```

### Methods

The package contains methods for plotting, printing and comparing spca solutions. These are:

- `choosecard`: interactive method for choosing the cardinality. It plots and prints statistics for comparing solutions of different cardinality.

- `print`: shows a formatted matrix of sparse loadings or *contributions* of a solution. Contributions are loadings expressed as percentages, while the loadings are scaled to unit sum of squares.

- `showload`: prints only the non-zero sparse loadings. This is useful when the number of variables is large.

- `summary`: shows formatted summary statistics of a solution

- `plot`: plots the cumulative variance explained by the sparse solutions versus that explained by the PCs, which is their upper bound. It can also plot the contributions in different ways.

- `compare`: plots and prints comparison of two or more *spca* objects.

### Minimal Example
The naming of the arguments in R is not simple, mainly because different syntaxes have been used over the years. I tried to give meaningful names starting differently so that R's useful feature of partial matching the arguments can be exploited. In the following wxample I sometime use partial arguments names.

```{r ex, fig.width = 3.5, fig.height = 3.5, fig.cap ="screeplot", fig.align= "center"}
library(spca)
cat(paste("loaded spca version:", packageVersion("spca")))
data(bsbl)

#- ordinary PCA
bpca = pca(bsbl, screeplot = TRUE, kaiser.print = TRUE)
```

```{r exsecond}
#- sparse PCA with minimal contribution 25%
bbe1 <- spcabe(bsbl, nd = 4, thresh = 0.25, unc = FALSE)

#- summary output
summary(bbe1)
#-# Explaining over 96% of the PCs' variance with 2, 3, 3 and 1 variables.

#- print percentage contributions
bbe1
#-# Simple combinations of offensive play in career and season are most important. Defensive play in season appears only in 3rd component.

#- The contributions can be printed one by one using the descriptive names in `bsbl_label`
data(bsbl_labels, package = "spca")
head(bsbl_labels)
showload(bbe1, variablesnames = bsbl_labels[,2])
#- plot solution
plot(bbe1, plotloadvsPC = TRUE, pc = bpca, mfr = 2, mfc = 2, 
               variablesnames = as.character(bsbl_labels[,2]))
#-# Explaining the variance pretty closely to PCA with much fewer variables.

#
```

### Installing the package

The package development is in the GitHub repository [GitHub repo](https://github.com/merolagio/spca)

* the latest released version from CRAN with

```{r instc, eval = FALSE}
install.packages("spca")
````

* The latest development version from github with

```{r instg, eval = FALSE}
if (packageVersion("spca") < 0.4.0) {
  install.packages("devtools")
}
devtools::install_github("merolagio/spca")
```

### Future releases
This is the first release and will surely contain some bugs, even though I tried to test it. Please do let me know if you find any or can suggest improvements. Please use the *Github* tools for submitting bugs [Bug report](https://github.com/merolagio/spca/issues/new ) or contributions.

For now most of the plots are produced with the basic plotting functions. In a later release I will produce the plots with ggplot2 (requires learning the package better).

The code is implemented in R, so it will not work for large datasets. 
I have in mind to develop C routines at least for the matrix algebra. Anybody willing to help, please, let me know. 

# References